---
title: "Wrangle_Data"
output: html_document
date: "2023-04-22"
---

This notebook walks you through wrangling the pilot data into cellphylo format.  
The end product of this notebook - UMI matrices in `cellphylo` format - are provided in the R object: xx  

# Preliminaries  
```{r setup}
#set working directory
setwd("~/repo/cellphylo/scripts/cellphylo_analysis/")

```

Create some directories  
```{r}

# for meta data annotation files
dir.create("./ann")
# for the  matrices created during cellphylo analysis
dir.create("./matrix")
# dir for raw csv files of count matrices 
dir.create("./matrix/raw_csv")

```

# Wrangle data  
We use the single cell atlas from [van Zyl et al.   (2020)](https://doi.org/10.1073/pnas.2001250117) as our pilot data.  

Download the meta data file provided by van Zyl et al. (2020):   [all_five_species_metafile.csv](https://singlecell.broadinstitute.org/single_cell/study/SCP780/).
This meta data file has information on sample ID, cell barcode and cluster ID (which corresponds to "cell type groups" in Mah & Dunn 2023).  
Deposit the meta data file in the `./ann` directory.  

## Wrangle the meta data file. 
```{r wrangle-metadata}

wrangled_meta <- wrangle_metadata(path_to_meta_file="./ann/all_five_species_metafile.csv")

```

This creates and deposites `all_five_species_metafile_wrangled.csv` in the `ann` directory. This file has parsed out the cell identifier (used by van Zyl as the cell names in the UMI matriecs), cluster_id, sample_id and cell_barcode.  

## Wrangle UMI matries  
Download the raw UMi matrices from NCBI GEO database (GSE146188).  

[human](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE148371)  
[macF (Macaca fascicularis)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE148373)  
[macM (Macaca mulatta)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE148374)  
[mouse](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE146186)  
[pig](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE146187)  

Deposit these gzipped files in `matrix/raw_csv`.  

### Convert csv files to 10x-style sparse matrices  

Convert the CSV files to 10x-style feature-barcode matrices. Sparse matrix format is easier to work with.  

```{r convert-to-10x}

#human
convert_to_10x(path_to_csv = "matrix/raw_csv/GSE148371_Human_count_matrix.csv.gz" ,  species_name = "human")
#macF
convert_to_10x(path_to_csv = "matrix/raw_csv/GSE148373_MacaF_count_matrix.csv.gz" ,  species_name = "macF")
#macM 
convert_to_10x(path_to_csv = "matrix/raw_csv/GSE148374_MacaqueM_count_matrix.csv.gz", species_name = "macM")
#mouse
convert_to_10x(path_to_csv = "matrix/raw_csv/GSE146186_Mouse_count_matrix.csv.gz" , species_name = "mouse")
#pig
convert_to_10x(path_to_csv = "matrix/raw_csv/GSE146187_Pig_count_matrix.csv.gz" , species_name = "pig")

```
Matrices are deposited in `matrix/10x/`.  

### Convert cell identifiers to cellphylo format  

Cell identifiers in the raw matrix provided by van Zyl et al.(2020) are currently:  
  <sample_id>_<cell_barcode>.  

Convert cell identifiers to `cellphylo` format.  

The cellphylo format is:  
  <species>_<sample_id>_<cluster_id>_<cell_type_id>_<cell_barcode>  

The fields are:  
  - species: species or animal name  
  - sample id: the id of the sequencing batch sample. Within-species integration will act on these batches.  
  - cluster id: aka 'cell type group' - the cluster id assigned by van Zyl et al. (2020). This consists of a cluster number (where present) and the cell type label.  
  - cell_type_id: the cell type label.  

This format is crucial as this information will be parsed repeatedly during cellphylo analysis from the cell names.  

```{r wrangle-matrix-names}

#convert cell identifiers to cellphylo format and create a new matrix with this new format
metadata_file = "ann/all_five_species_metafile_wrangled.csv" 


#human
wrangled_mat_human <- wrangle_matrix(path_to_mat = "matrix/10x/matrix_10x_human", metadata_file = metadata_file, species = "human")
#macF
wrangled_mat_macF <- wrangle_matrix(path_to_mat = "matrix/10x/matrix_10x_macF", metadata_file = metadata_file, species = "macF")
#macM
wrangled_mat_macM <- wrangle_matrix(path_to_mat = "matrix/10x/matrix_10x_macM", metadata_file = metadata_file, species = "macM")
#mouse
wrangled_mat_mouse <- wrangle_matrix(path_to_mat = "matrix/10x/matrix_10x_mouse", metadata_file = metadata_file, species = "mouse")
#pig
wrangled_mat_pig <- wrangle_matrix(path_to_mat = "matrix/10x/matrix_10x_pig", metadata_file = metadata_file, species = "pig")


#check that cell ids are in cell phylo format
colnames(wrangled_mat)

```


Congrats! You have matrices in cellphylo format. :)  

These wrangled matrices have been provided in the R object: xxx.  

