---
title: "Create Base Matrix"
output: html_notebook
---

Run `1_wrangle_data.Rmd` to wrangle UMI matrices into cellphylo format (Fig. S1, step 1).

This notebook walks you through Fig. S1, steps 2-7, of the cellphylo workflow created for Mah & Dunn 2023. 

The steps are:  
2 & 3. Normalization and within-species integration
4. Randomly subset matrices
5. Combine matrices by gene orthology 
6. Cross-species integration  
7. PCA: rotation and rank reduction of data  

This will produe a 919 cell 919 PC matrix from which the experiments of this paper were performed upon. This matrix is stored in the R object: xxx.  


# Preliminaries
```{r setup}
#set working directory
setwd("~/repo/cellphylo/scripts/cellphylo_analysis/")

#load libraries
library(tidyverse)
library(Seurat)
library(DropletUtils)
```


# Within-Species Analysis
The within-species analysis is performed separately for each species. 

## Steps 2 & 3. Normalization and Within-species Integration  
Fig. S1 steps 2 and 3

For each species, we use the Seurat workflow to normalize counts using a negative binomial model (via SCTransform) and integrate sequencing batches within each species.    

This step is memory intensive and you may have to run this on an HPC cluster.  
The human data set required 125 GiB RAM per CPU.  
All other species' data sets required 50 Gib ram per CPU.

Parameter values used by Mah & Dunn (2023) for `within_species_integration` are set as default.  

```{r within-species-integration}

#human
human_obj <- within_species_integration(path_to_matrix = "matrix/wrangled/matrix_wrangled_humanF", n_features=3000, k_weight = 100, print=TRUE)

#macF
macF_obj <- within_species_integration(path_to_matrix = "matrix/wrangled/matrix_wrangled_macF", n_features=3000, k_weight = 100, print=TRUE)

#macM
macM_obj <- within_species_integration(path_to_matrix = "matrix/wrangled/matrix_wrangled_macM", n_features=3000, k_weight = 100, print=TRUE)

#mouse
mouse_obj <- within_species_integration(path_to_matrix = "matrix/wrangled/matrix_wrangled_mouse", n_features=3000, k_weight = 100, print=TRUE)

#pig
pig_obj <- within_species_integration(path_to_matrix = "matrix/wrangled/matrix_wrangled_pigF", n_features=3000, k_weight = 100, print=TRUE)

```

Moving forward, we want to use the matrix of integrated Pearson's residuals, labelled: `matrix_sct_integrated_pearson_residuals`.  

## Step 4. Randomly subset matrices  
Fig. S1 step 4

### Identify the cell type with the fewest replicates
To make later analyses computationally feasible, we randomly subset each species' matrix so that all cell type groups across all five species have the same number of replicates. This means subsetting to the smallest number of replicates displayed by any one cell type group.  

```{r count-replicates}

count_replicates("matrix/within-species")

```

We determine that the cell type group with the smallest number of replicates are mouse B cells, with 10 cells.  

### Subset the matrices 

For each species, randomly subset the full matrix to 10 cells per cell type group.  
```{r subset-cells} 
#human
subset_cells(path_to_matrix = "matrix/within-species/matrix_sct_integrated_pearson_residuals_human", n_reps = 10, print=TRUE)
#macF
subset_cells(path_to_matrix = "matrix/within-species/matrix_sct_integrated_pearson_residuals_macF", n_reps = 10, print=TRUE)
#macM
subset_cells(path_to_matrix = "matrix/within-species/matrix_sct_integrated_pearson_residuals_macM", n_reps = 10, print=TRUE)
#mouse
subset_cells(path_to_matrix = "matrix/within-species/matrix_sct_integrated_pearson_residuals_mouse", n_reps = 10, print=TRUE)
#pig
subset_cells(path_to_matrix = "matrix/within-species/matrix_sct_integrated_pearson_residuals_pig", n_reps = 10, print=TRUE)

```

# Cross-species Analysis
The cross-species analysis is performed on a combined matrix of all five species (Fig. S1, steps 5-7).

## Step 5. Combine matrices by gene orthology 
Fig. S1 step 5

We will combine the species matrix into a single matrix based on shared orthlogs.  

### Wrangle the orthlogs table  

Gene symbols vary between species. We need to create a mapping between gene symbols before we can combine the species' matrices by gene orthology. We will obtain this mapping from [Ensembl Biomart](https://www.ensembl.org/biomart/martview/).  

The below code is designed to data wrangle the Ensembl ortholog file downloaded for Mah and Dunn (2023). This wrangling facilitates downstream analysis. I have provided this orthology table: `Aqhumor_ensembl_orthologs.txt`. Put this orthology table in the `ann` folder. 


```{r wrangle-ortholog-map}
#library(tidyverse)

# read in ensembl ontology file
ensembl_map <- read_tsv("ann/AqHumor_ensembl_orthologs.txt")
# standardize column names manually
colnames(ensembl_map) <- c("human_genename", "macM_genename","macM_homologytype", "macF_genename", "macF_homologytype","mouse_genename","mouse_homologytype", "pig_genename", "pig_homologytype")
#remove NAs
ensembl_map_noNA <- na.omit(ensembl_map)
#filter for lines where every seq is a one2one ortholog
ensembl_map_one2one <- filter(ensembl_map_noNA, macM_homologytype == "ortholog_one2one" & macF_homologytype == "ortholog_one2one" & mouse_homologytype == "ortholog_one2one" & pig_homologytype == "ortholog_one2one")
#add gene identifiers
ensembl_map_final <- mutate(ensembl_map_one2one, gene_identifier=paste0("gene_",1:nrow(ensembl_map_one2one)))
#select just gene name columns
ensembl_map_final <- select(ensembl_map_final, gene_identifier, macM_genename, macF_genename, mouse_genename, pig_genename, human_genename)
colnames(ensembl_map_final) <- c("gene_identifier", "macM", "macF", "mouse", "pig", "human")
                         

write.table(ensembl_map_final, file = "ann/AqHumor_ensembl_orthologs_wrangled.tsv", quote=FALSE, sep="\t", col.names=TRUE, row.names=FALSE)

```

This produces `AqHumor_ensembl_orthologs_wrangled.tsv` and TSV table mapping gene symbols across species. We use the generic gene_identifier (gene-1, gene-2...) moving forward.

### Combine all five species' matrices by shared orthologs

Combine the subsetted matrices by shared orthlogs.  

```{r combine-matrices}

matrix_dir_path <- "matrix/subset/"

#pursue old order for now
matrix_list <- c(paste0(matrix_dir_path,"matrix_subset_10_rep_macM"),paste0(matrix_dir_path,"matrix_subset_10_rep_macF/"),paste0(matrix_dir_path,"matrix_subset_10_rep_mouse"),paste0(matrix_dir_path,"matrix_subset_10_rep_pig/"),paste0(matrix_dir_path,"matrix_subset_10_rep_human/"))

#matrix_list <- c(paste0(matrix_dir_path,"matrix_subset_10_rep_human"),paste0(matrix_dir_path,"matrix_subset_10_rep_macF/"),paste0(matrix_dir_path,"matrix_subset_10_rep_macM"),paste0(matrix_dir_path,"matrix_subset_10_rep_mouse/"),paste0(matrix_dir_path,"matrix_subset_10_rep_pig/"))


combined_matrix <- combine_matrices(matrix_list = matrix_list, ortholog_map_path = "ann/AqHumor_ensembl_orthologs_wrangled.tsv", print=TRUE)

```

This produces the matrix `matrix/combined_by_orthology/matrix_combined` combined matrix is not yet cross-species integrated.

## Step 6. Cross-Species integration
Fig. S1 step 6

Use the Seurat workflow to integrate counts across species. Default values for cellphylo::cross_species_integration are what was used for Mah & Dunn (2023).

```{r cross-species-integration}

cross_species_int <- cross_species_integration(matrix_path = "matrix/combined_by_orthology/matrix_combined")

#let's visualize the cross-species integration
#check by species integrated?
p1 <- DimPlot(cross_species_int, reduction = "umap", group.by = "species_id")
p1
#check by cell types separate?
p2 <- DimPlot(cross_species_int, reduction = "umap", group.by = "cell_type_id")
p2
#cell types in each species present in the same clusters across species?
p3 <- DimPlot(cross_species_int, reduction = "umap", group.by = "cell_type_id", split.by = "species_id")
p3
```
This produces the matrix `mtrix/cross-species_integration/matrix_cross-species_integrated`.

#7. PCA: rotation and rank reduction of the data 
Fig. S1 step 7

## Inspect the distribution of variance across principal components
```{r plot-eigenvalues}
eig_plots <- plot_eigs(matrix_path = "matrix/cross-species_integration/matrix_cross-species_integrated/", highlight=20, print=TRUE)

```

This produces four plots:
1_eigenvalue_vs_PC: plot of eigenvalues against PCs. PC 20 is indicated with a red dot for Mah & Dunn (2023)
2_eigenvalue_cumulative_sum: cumulative sum of eigenvalues plotted against PCs
3_eigenvalue_vs_variance: plot of eigenvalue vs percentage variance described by each PC.
4_eigenvalue plot of percenrtage variance described by each PC against PCs. PC 20 is indicated wiht a red dot for Mah & Dunn (2023.)

From these plots we can identify an elbow at PC20. 

## Use PCA to rotate and rank reduce the matrix

Perform PCA on the matrix. Mean-centre before PCA and normalize variance by standard deviation. 
```{r run-pca}
pca_matrix <- run_pca(matrix_path="matrix/cross-species_integration/matrix_cross-species_integrated", n_PCs=919, print=TRUE)
```


Congratulations! You have created the 919 cell 919 PC matrix from Fig. S1 step 7. ALl other experiments and figures are based on this matrix.

Because sampling is random, we have provided the 919 cell 919 PC matrix used in our analysis: 


# Create matrices for Figure 4.  

## Create 919 cell 20 PC matrix (Fig. S1, step C1)

```{r run-pca-20pc}
pca_matrix <- run_pca(matrix_path="matrix/cross-species_integration/matrix_cross-species_integrated", n_PCs=20, print=TRUE)
```


## Create 92 cell matrix (Fig. S1, step C2)
Preliminary `contml` runs were performed on a 92 cell 20 PC matrix (Fig. S1, step C2), which contains 1 cell per cell type group per species. All cell type groups (92) were retained. 

Because sampling is random, we have provided the 92 cell matrix we used in our analysis:`contml_92cell_subset_pca_var_norm_20PC_mtx/`. However, if you'd like to create a new 92 cell matrix use the below code:

```{r create-92cell-mat}
#deposit matrix in folder
if(!dir.exists("figure_4")){
  dir.create("figure_4")
}

if(!dir.exists("figure_4/matrix")){
  dir.create("figure_4/matrix")
}

#this prints out a list of randomly subsetted cells, 1 cell per cell type group per species
subset_mat <- subset_combined_matrix(matrix_path = "matrix/cross-species_integration/matrix_cross-species_integrated", sample_size=1, print=TRUE)
#This produces a list of cells that have been subsetted from the cross-species integrated matrix

file.rename("matrix_combined_92cell_subset_mtx", "figure_4/matrix/matrix_combined_92cell_subset_mtx")
file.rename("matrix_combined_92cell_subset.txt", "figure_4/matrix/matrix_combined_92cell_subset.txt")

#The list that is produced is then used to subset the 919 cell 919 PC matrix. We can use the run_pca() command to do so.
subset_mat_pca <- run_pca(matrix_path = "matrix/cross-species_integration/matrix_cross-species_integrated", n_PCs = 20, subset_file_path = "figure_4/matrix//matrix_combined_92cell_subset.txt", print=TRUE)

file.rename("matrix/PCA/contml_92cell_subset_20PC_mtx", "figure_4/matrix/contml_92cell_subset_20PC_mtx")

```

## Create 54 cell matrix (Fig. S1, step C3)
We further subsetted the 92 cell matrix (Fig. S1, step C2) to 54 cells (Fig. S1, step C3).

The following script intakes the 92 cell 20 PC matrix (Fig. S1, step x) and remove cell type groups that either: 1) consistently failed to form clades and disrupted other cell type clades, or 2) existed only in a single species. 

```{r create-54cell-mat}

#for reproducibility, use the 92 cell matrix provided by Mah & Dunn (2023)
mat <- Read10X("contml_92cell_subset_pca_var_norm_20PC_mtx")

#remove unstable cells or singletons
black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal", "Endo", "Endo_Schlemms", "Epi-CiliaryNonPigment", "Epi-CiliaryPigment","Epi-Pigment", "MastCell", "Myoepithelium", "Neuron", "SchwalbeLine", "Uveal","Vascular","Beam-A", "Beam-X", "Epi-Corneal", "Fibroblast", "BCell")


cell_ids <- rownames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

filter_index <- which(!(cell_type_ids %in% black_list))

mat_filt <- mat[filter_index,]

#write matrix
#write10xCounts("contml_57cell_subset_pca_var_norm_20PC_mtx", mat_filt, version="3")
write10xCounts("contml_54cell_subset_20PC_mtx", mat_filt, version="3")

file.rename("contml_54cell_subset_20PC_mtx", "figure_4/matrix/contml_54cell_subset_20PC_mtx")

```

This is the matrix used to create the tree in Figure 4. 

We provide this matrix and infile as:
`contml_54cell_subset_pca_var_norm_20PC_mtx`
`infile_54_20PC`  

## Create 540 cell matrix (Fig. S1, step C4)
This matrix is used for the single cell jackknife analysis that produced the scjackknife scores in Fig. 4.

We create it by removing unstable cells and singletons from the 919 cell cross-species integrated matrix. This leaves 54 cell type groups, each with 10 cells each. This matrix contains 2000 genes.

```{r create-540cell-mat}

mat <- Read10X("matrix/cross-species_integration/matrix_cross-species_integrated/")

black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal", "Endo", "Endo_Schlemms", "Epi-CiliaryNonPigment", "Epi-CiliaryPigment","Epi-Pigment", "MastCell", "Myoepithelium", "Neuron", "SchwalbeLine", "Uveal","Vascular","Beam-A", "Beam-X", "Epi-Corneal", "Fibroblast", "BCell")


cell_ids <- colnames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

filter_index <- which(!(cell_type_ids %in% black_list))

mat_filt <- mat[,filter_index]

#write matrix
write10xCounts("matrix_540cell_cross_species_integrated_mtx", mat_filt, version="3")
file.rename("matrix_540cell_cross_species_integrated_mtx", "figure_4/matrix/matrix_540cell_cross_species_integrated_mtx")
```
input:
aqhumor_cross_species_integrated_mtx mtx md5: daa2aef295e2b7f07f80de669b8a3c1c

output:
540_aqhumor_cross_species_integrated_mtx mtx md5: d616301cc9fdda896ed4eac52268f64f


