---
title: "3_PC_sweep_exp"
output: html_document
date: "2023-04-25"
---

This notebook outlines how to reproduce the PC sweep experiment (Fig. 2, Mah & Dunn 2023).
To perform the PC sweep experiment, we perform PCA on the cross-species integrated matrix (Fig. S1, step 6) to create a 919 cell 919 PC matrix (Fig. S1 step 7). We then subset it from 3 to 100 PCs, producing 98 matrices which we converted into `contml` infiles for phylogenetic inference. 

# Preliminaries

```{r}
#set working directory to the analysis directory
setwd("~/repo/cellphylo/analysis")

```


# Make PC sweep infiles
Subset the 919 cell 919 PC matrix to 3-100 PCs and print out the matrices as `contml` infiles.

```{r make-pc-sweep-infiles}

#perform PCA on the cross-species integrated matrix.
make_pc_sweep_infiles(matrix_path="matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/", run_PCA=TRUE, PC_range=c(3:10))

```

To facilitate reproducibility, we have provided the infiles in `PC_sweep/input/pc_sweep_infiles_final.tar.gz`*


# Infer a cell phylogeny from each infile with `contml`  
For each PC sweep infile, infer a tree using `contml`.

`contml` is part of PHYLIP. Install [`PHYLIP v. 3.697`](https://github.com/dunnlab/phylip). We ran PHYLIP in an [Ubuntu docker container](https://github.com/dunnlab/ubuntu-dev).

`contml` was run using the 'C' (continuous) option, and all other options left at default.  
The bash script `scripts/PC_sweep/contml_pc_sweep_919.sh` was used to infer the trees for the pc sweep analysis. 

*The inferred trees are provided in `PC_sweep/output/pc_sweep_919_outtrees.tar.gz`*

# Calculate and plot the PC sweep variables
We will calculate total tree length (sum of all branch lengths), the sum of interior edge lengths, sum of tip edge lengths, and star-ness score (ratio of tip to interior edge lengts) from the `outtrees` produced by `contml`.

```{r calculate-plot-PC-sweep}

#calculate the PC sweep variables. 
variables.df <- calc_pc_sweep_var("PC_sweep/output/pc_sweep_919_outtrees/")

#plot the variables to create the graphs in Fig. 2
plots <- plot_pc_sweep(rank=c(3:100), variables.df, print=TRUE)

```



