---
title: "Jumble trees"
---

# Jumble tree analysis
The following workflow was  used to estimate 'jumble trees' - replicate trees inferred from a single data set by jumbling the order of input taxa to contml. The clade frequency across these jumble trees were used as "jumble scores", a measure of technical repeatability. 

# Setup
```{r setup}
library(Seurat)
library(ape)
library(phangorn)
library(proxy)

#set the working directory to the analysis directory
setwd("~/repo/cellphylo/analysis")

```

# Create matrices for inference of the focal phylogeny in Figure 4

## Create 92 cell 20 PC matrix (Fig. S1, step C1.1)
Preliminary `contml` runs were performed on a 92 cell 20 PC matrix (Fig. S1, step C2), which contains 1 cell per cell type group per species. All cell type groups (92) were retained. 

Because sampling is random, for reproducibility we have provided the 92 cell matrix and list of sampled cells we used in our analysis: *`tree_inference/92_cell_tree/contml_92cell_subset_pca_var_norm_20PC_mtx/`*.

However, if you'd like to create a new 92 cell matrix use the below code:

```{r create-92cell-mat}
#deposit matrix in folder
if(!dir.exists("tree_inference")){
  dir.create("tree_inference")
}

if(!dir.exists("tree_inference/92_cell_tree")){
  dir.create("tree_inference/92_cell_tree")
}

#this prints out a list of randomly subsetted cells, 1 cell per cell type group per species. These cells are drawn from the cross-species integrated matrix
subset_mat <- subset_combined_matrix(matrix_path = "matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/", sample_size=1, print=TRUE)

file.rename("aqhumor_cross_species_92_subset_mtx", "tree_inference/92_cell_tree/aqhumor_cross_species_92_subset_mtx")

file.rename("aqhumor_cross_species_92_subset.txt", "tree_inference/92_cell_tree/aqhumor_cross_species_92_subset.txt")

#The list that is produced is then used to subset the 919 cell 919 PC matrix. We can use the run_pca() command to do so.
subset_mat_pca <- run_pca(matrix_path = "matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/", n_PCs = 20, subset_file_path = "tree_inference/92_cell_tree/aqhumor_cross_species_92_subset.txt", print=TRUE)

file.rename("contml_92cell_subset_pca_var_norm_20PC_mtx", "tree_inference/92_cell_tree/contml_92cell_subset_pca_var_norm_20PC_mtx")

```

## Create 54 cell 20 PC matrix (Fig. S1, step C1.2)
We further subsetted the 92 cell matrix (Fig. S1, step C1.1) to 54 cells (Fig. S1, step C1.2).

The following script intakes the 92 cell 20 PC matrix (Fig. S1, step C1.1) and remove cell type groups that either: 1) consistently failed to form clades and disrupted other cell type clades, or 2) existed only in a single species. 


```{r create-54cell-mat}

#for reproducibility, we have provided the 92 cell 20 PC matrix
mat <- Read10X("tree_inference/92_cell_tree/contml_92cell_subset_pca_var_norm_20PC_mtx/")

#remove unstable cells or singletons
black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal", "Endo", "Endo_Schlemms", "Epi-CiliaryNonPigment", "Epi-CiliaryPigment","Epi-Pigment", "MastCell", "Myoepithelium", "Neuron", "SchwalbeLine", "Uveal","Vascular","Beam-A", "Beam-X", "Epi-Corneal", "Fibroblast", "BCell")


cell_ids <- rownames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

filter_index <- which(!(cell_type_ids %in% black_list))

mat_filt <- mat[filter_index,]

#write matrix

write10xCounts("contml_54cell_subset_pca_var_norm_20PC_mtx", mat_filt, version="3")

if(!dir.exists("tree_inference")){
   dir.create("tree_inference")}

if(!dir.exists("tree_inference/54_cell_tree")){
    dir.create("tree_inference/54_cell_tree")
    }
file.rename("contml_54cell_subset_pca_var_norm_20PC_mtx", "tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx")

```

For reproducibility we provide the 54 cell matrix we used in our analyses in:  *`tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx`*.

This is the matrix used to create the focal phylogeny in Figure 4. 

# Jumble analysis

This workflow calculates jumble trees, removes errored files, and then calculates jumble scores.

## Create infile for jumble tree

The jumble analysis was performed on the 54 cell 20 PC matrix (Fig. S1, step C1.2). This matrix has been provided in `tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx/` for reproducibility.  

This matrix must be converted into an `infile`, the input file format `contml` requires.

```{r}

make_infile(matrix_path = "tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx/", print=TRUE)

if(!dir.exists("tree_inference")){
   dir.create("tree_inference")}

if(!dir.exists("tree_inference/54_cell_tree")){
    dir.create("tree_inference/54_cell_tree")
    }
file.rename("infile", "tree_inference/54_cell_tree/infile")


```
This infile has been provided in *`jumble/input/infile_54_20PC`*, for convenience. `contml` expects the input file to be named "infile", so rename this file when running.


## Infer jumble trees from contml runs

The bash script *`script_jumble_parallel_54.sh`* was used to infer the jumble trees from the infile. This script produces a random seed, tells `contml` to randomly jumble input order 100x, and performs a maximum likelihood tree search to produce 100 jumble trees. From these trees, the jumble tree with the highest likelihood is selected as the final output of the run. This was performed 250 times to obtain 250 high-likelihood jumble trees for downstream analysis. 

The script *`run_jumble_parallel_54.sh`* was used to run `script_jumble_parallel_54.sh` in parallel on a hpc cluster.

The `contml` settings:
  C: continuous characters
  J:   Randomize input order of species (jumble)
    Random number to seed (see bash script)
    Number of times to jumble: 100
  G: Global rearrangements
  
The bash scripts are available in `scripts/jumble/`.


## Remove missing/errored files
Some Jumble runs may error and produce no tree. We must remove those files.

Delete any empty tree files using the bash command.
```
find ./ -size 0 -print -delete
```

Use the `outtree` directory with empty tree files removed for all downstream analysis. This directory has been provided as *`jumble/output/outtrees_54_jumble_search_processed.tar.gz`*


## Calculate jumble scores and identify the best-scoring jumble tree

### Identify best jumble tree

```{r id-best-jumble-tree}

if(!dir.exists("jumble")){
    dir.create("jumble")
}

#path to contml outtrees (empty files removed)
tree_dir_path = "jumble/output/outtrees_54_jumble_search_processed/"

#find best jumble tree
best_jumble_tree(tree_dir_path = tree_dir_path)
list.files(tree_dir_path)

```

14 jumble trees tied for best sum of scores.

They are:
 index 18: outtree_54_jumble_search_1235256431.tre
 index 20: outtree_54_jumble_search_1249177875.tre
 index 56: outtree_54_jumble_search_1790537341.tre
 index 82: outtree_54_jumble_search_2214045959.tre
 index 94: outtree_54_jumble_search_2481088323.tre
 index 103: outtree_54_jumble_search_2642665967.tre
 index 123: outtree_54_jumble_search_2892552345.tre
 index 149 :outtree_54_jumble_search_3280076499.tr
 index 169: outtree_54_jumble_search_3567552921.tre
 index 209: outtree_54_jumble_search_542946467.tre
 index 211: outtree_54_jumble_search_545341417.tre
 index 217: outtree_54_jumble_search_61674183.tre
 index 223: outtree_54_jumble_search_683352147.tre
 index 234: outtree_54_jumble_search_909119069.tre


### Label and score each best jumble tree

```{r label-score-jumbles}

#path to contml outtrees (empty files removed)
tree_dir_path = "jumble/output/outtrees_54_jumble_search_processed/"
#combine trees into a single multiphylo object
multiphylo <- combine_multi_trees(tree_dir_path)

#read in input matrix that was used to create the infile contml used to infer jumble trees
matrix_path <- "tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx/"
mat <- Read10X(matrix_path)

#list of the 14 best scoring jumble trees
tree_list <- c("outtree_54_jumble_search_1235256431.tre", "outtree_54_jumble_search_1249177875.tre", "outtree_54_jumble_search_1790537341.tre","outtree_54_jumble_search_2214045959.tre", "outtree_54_jumble_search_2481088323.tre","outtree_54_jumble_search_2642665967.tre","outtree_54_jumble_search_2892552345.tre","outtree_54_jumble_search_3280076499.tre",  "outtree_54_jumble_search_3567552921.tre", "outtree_54_jumble_search_542946467.tre", "outtree_54_jumble_search_545341417.tre","outtree_54_jumble_search_61674183.tre", "outtree_54_jumble_search_683352147.tre", "outtree_54_jumble_search_909119069.tre") %>% as.list()

#create a directory to deposit labelled trees
dir.create("jumble/best_trees")

#label tips and jumble scores of the best-scoring jumble trees.
labelled_trees <- lapply(tree_list, function(trees){
  focal_tree_path <- paste0(tree_dir_path,trees)
  
  #score tree
  tree <- score_jumble(multiphylo = multiphylo, focal_tree_path = focal_tree_path, print=FALSE)

  #label tree
  #connect cell ids to index 
  key.df <- data.frame(cell_id = rownames(mat), index = as.character(c(1:nrow(mat))))

  #find tip label
  tips.df <- data.frame(index = tree$scored$tip.label)

  #connect tip id to cell id
  new_tip.df <- left_join(tips.df, key.df, by="index")

  #replace old tip indices with new tip cell id labels
  tree$scored$tip.label <- new_tip.df$cell_id
  
  file_name = trees
  
  #write the labelled trees out to file
  write.tree(tree$scored, file=paste0("jumble/best_trees/",file_name))
  
  return(tree)
})


```


### Choose the most representative 'best tree'
There are 14 best trees tied for best sum/mean of jumble scores. Choose the tree that is most representative of the best trees. We identify this tree by calculating the distances between the 14 trees among each other and selecting the tree with the smallest sum of distances. I examined Robinson-Foulds, path-distance and Kuhner-Felsenstein distance. 

```{r}

#create a multiphylo of best trees
tree_dir_path <- "jumble/best_trees/"
multiphylo <- combine_multi_trees(tree_dir_path)

#make a distance matrix. See phangorn::treedist for description of distance metrics.
#using Robinson-Foulds distance
RF_dist_mtx <- RF.dist(multiphylo) #all distances 0
RF_dist_mtx

#using path distance
path_dist_mtx <- path.dist(multiphylo) #all distances 0
path_dist_mtx 

#using Kuhner & Felsenstein (1994) distance.
KF_dist_mtx <- KF.dist(multiphylo) #distances vary when branch length is taken into account
KF_dist_mtx

#which tree has the smallest sum of KF_dists?
KF_mat <- as.matrix(KF_dist_mtx)
colsums <- colSums(KF_mat)
min <- min(colsums) #0.2320657

#Tree index 2 < tree index 11

```

The tree with the minimum *unrounded* Kuhner-Felsenstein distance with all other "best trees" is `outtree_54_jumble_search_1249177875.tre`. Thus, this tree and its jumble scores was chosen to provide the topology of the tree in Figure 4. 

We provide this scored tree at *`analysis/jumble/output/outtree_54_jumble_search_1249177875_scored_label.tre`.*

