---
title: "single cell jackknife analysis"
---

This notebook provides the workflow to calculate the single cell jackknife (scjackknife) scores for Figure 4. The scjackknife procedure randomly draws replicate cells from a larger matrix to create the 54 cell matrix. We repeat this procedure 500 times and infer an `scjackknife tree` from each. These trees are used to calculate TBE (transfer bootstrap expectation) scores, which served as the `scjackknife scores` in Figure 4.

#Preliminaries
```{r}
setwd("~/repo/cellphylo/analysis")

#load 
library(ape)

#create directory

if(!dir.exists("scjackknife")){
  dir.create("scjackknife")
}


```

# scJackknife analysis
This workflow creates scjackknife matrices, infers a scjackknife tree from each, removes errored runs and calculates TBE scores from the scjackknife trees. 

#Create the matrices we will sample from

## Create 919 cell 20 PC matrix

```{r run-pca-20pc}
pca_matrix <- run_pca(matrix_path="matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/", n_PCs=20, print=TRUE)
```


*To facilitate reproducibility, we have provided this matrix in `scjackknife/contml_919cell_subset_20PC_mtx`.*


## Create 540 cell matrix to draw cell labels from (Fig. S1, step C2.2)
We create it by removing unstable cells and singletons from the 919 cell cross-species integrated matrix. This leaves 54 cell type groups, each with 10 cells each. We only sample the cell labels from this matrix, not the values, and the cell labels we sample from this matrix will be used to subset the 919 cell 20 PC matrix that was created above.

```{r create-540cell-mat}

mat <- Read10X("matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/")

black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal", "Endo", "Endo_Schlemms", "Epi-CiliaryNonPigment", "Epi-CiliaryPigment","Epi-Pigment", "MastCell", "Myoepithelium", "Neuron", "SchwalbeLine", "Uveal","Vascular","Beam-A", "Beam-X", "Epi-Corneal", "Fibroblast", "BCell")


cell_ids <- colnames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

filter_index <- which(!(cell_type_ids %in% black_list))

mat_filt <- mat[,filter_index]

#write matrix
write10xCounts("matrix_540cell_cross_species_integrated_mtx", mat_filt, version="3")
file.rename("matrix_540cell_cross_species_integrated_mtx", "scjackknife/matrix_540cell_cross_species_integrated_mtx")
```

*We provide this matrix in `scjackknife/matrix_540cell_cross_species_integrated_mtx`*.

# Make scjack samples
To create each replicate 54 cell scjack matrix, we will repeatably sample 1 cell per cell type group per species, for the 54 cell type groups we used to create the phylogeny in Figure 4. We will use a 540 cell matrix (10 cells per cell type group, 54 cell type groups) to draw cell labels from. Because PCA must be performed on the full 919 cell matrix, we will create a 919 cell 20 PC matrix (Fig. S1, step C1) and pull out the values corresponding to the cell labels we obtained from subsampling the 450 cell matrix. 

This in effect is the same as subsetting the 919 cell 20 PC matrix to 450 cell and 20 PCs (Fig. S1, step C4), but due to a quirk in the workflow the former was performed.

We do this 500 times to create 500 resampled 54 cell matrices. Each matrix is transformed into a `contml infile` and contml is used to infer  tree from it using the same settings as that used for the jumble trees.

This function will print out the 54 cell subsampled matrices as 10x feature-barcode matrices, infiles, and the list of cell ids present in each matrix.

*Since the scjackknife matrices are random, we have included the input we used as `scjackknife/input_scjack_540_TBE.tar.gz`*

```{r create-scjack-sample}

# The matrix to subsample cells from. This is the 540 cell matrix (Fig. S1, step C4)
subsample_matrix_path = "scjackknife/matrix_540cell_cross_species_integrated_mtx/"
# The matrix to draw values from. This should be the 919 cell 20 PC matrix (Fig. S1, step C1)
full_matrix_path = "scjackknife/contml_919cell_subset_20PC_mtx/"
#number of scjack samples to make
n_samples = c(1:10)

#function calls: subset_combined_matrix(), run_pca(), and make_infile()
scjack_sample <- make_scjack_sample(n_samples = n_samples, subsample_matrix_path = subsample_matrix_path, full_matrix_path = full_matrix_path, file_name="540_TBE")

```

# Create scripts to infer scjackknife trees with contml
Use the infiles created above to run `contml`.

`contml` was run using the same options as for the jumble trees:
  C: continuous data
  J: jumble, 100x with random seed
  G: global search

We ran this on a hpc cluster using 64 vCPUs and running 50 `contml` commands in parallel at a time. This took 4-5 hours.

Each scjack `contml` run requires its own bash script. We created and executed 500 bash scripts using the following workflow:

1. We created a template for the script that is executed for each contml run. This has been provided as `analysis/scripts/scjackknife/in_script_scjack_parallel_540_TBE.sh`

2. We made 500 copies of the template script. The script for this is provided as `analysis/scripts/scjackknife/make_reps_scjack_540_TBE.sh`.

3. Execute each scjackknife script in parallel. The script for this is `analysis/scripts/scjackknife/run_scjack_parallel_540_TBE.sh`.

*We have included the resulting scripts in `scripts/scjackknife/scjack_scripts.tar.gz`*

This produced 500 outtrees representing a scjackknife tree inferred from each resampled 54 cell matrix.

# Process output files
Some `contml` runs may have errored out, producing empty `outtree` files. Remove these files

Delete any empty tree files
```
find ./ -size 0 -print -delete
```

There are 439 scjackknife trees remaining.

Moving forward, use the directories of `outtrees` that have had errored runs removed for downstream analyses. *This directory has been provided as `scjackknife/output/output_processed/outtrees_scjack_540_TBE_processed`.*

Removing files causes the index between the scjack matrices and `outtree` and matrix files to be mismatched, causing downstream analysis issues. Remove those scjack matrices that produced errored runs.

```{r id-errored-runs}
outtrees_dir = "scjackknife/output/raw/outtrees_scjack_540_TBE/"
matrix_name_pattern = "scjack_540_TBE"

id_errored_runs(outtrees_dir = outtrees_dir, matrix_name_pattern = matrix_name_pattern)

```

The above function provides a list of matrices whose `contml` runs errored. Use this list to remove these matrices using bash:
```
cd analysis/scjackknife/input_scjack_540_TBE/scjack_matrices
rm -r $(<errored_runs.txt)

```

Moving forward, use the directory of matrices with error-causing matrices removed in downstream analysis. *This has been provided as `scjackknife/output/output_scjack_540_TBE_processed/scjack_matrices_processed`*

# Prepare input files for Booster.

## Add common labels to scjackknife trees
scjackknife trees must share common labels in order to calculate the TBE. Label the scjackknife trees with their full cell ids and an abbreviated cell id that can be matched across scjackknife trees.


```{r prep-scjack-boost}

n_trees =439
matrix_dir_path = "scjackknife/output/output_processed/scjack_matrices_processed/"
tree_dir_path = "scjackknife/output/output_processed/outtrees_scjack_540_TBE_processed/"
file_name_prefix = "outtrees_scjack_540_TBE"

labelled_trees <- label_scjack_trees(n_trees = n_trees, matrix_dir_path = matrix_dir_path, tree_dir_path = tree_dir_path, print=TRUE, file_name_prefix=file_name_prefix)

```

The trees that have been labelled with common labels have been provided in 
*`scjackknife/output/output_rocessed/outtrees_scjack_540_TBE_common_label`*

## Add common labels to focal tree

The focal tree must also have the common labels
```{r prep-focal-boost}

#The phylogeny to calculate TBE scores for must also share common labels with the scjackknife trees.
#below is the tree used for Fig. 4
tree <- read.tree("jumble/output/outtrees_54_jumble_search_processed/outtree_54_jumble_search_1249177875.tre")

labelled_tree <- label_tree(matrix_path = "tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx/", tree_path = "jumble/output/outtrees_54_jumble_search_processed/outtree_54_jumble_search_1249177875.tre", file_name = "outtree_54_jumble_search_1249177875_label.tre", print=FALSE)

#make new labels - code stolen from scjack label
species <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[3])})
new_label <- paste0(species, "_", cluster_id)
common_labels_tree <- labelled_tree
common_labels_tree$tip.label <- new_label
#print out common labels tree
write.tree(common_labels_tree, file="outtree_54_jumble_search_1249177875_common_label.tre")



```

*To facilitate reproducibility, this tree has been provided in `scjackknife/booster/outtree_54_jumble_search_1249177875_common_label.tre`*

#Combine common label scjackknife trees in a single multiphylo file
```{r boost-files}

#combine and print out as a single tree file
multi_trees <- combine_multi_trees("scjackknife/output/output_scjack_540_TBE_processed/outtrees_scjack_540_TBE_common_label/")

write.tree(multi_trees, file="multiphylo_scjack_540_TBE_common_label.tre")

    
```

*To facilitate reproducibility, this multi tree file has been provided in `scjackknife/booster/multiphylo_scjack_540_TBE_common_label.tre`*

## Run booster to calculate TBE scores
```
./booster/booster_macos64 -i outtree_54_jumble_search_1249177875_common_label.tre -b multiphylo_scjack_540_TBE_common_label.tre  > outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre
```

This produces a tree, `outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre`, with TBE scores plotted at nodes. These TBE scores were used in Fig. 4 as the scjackknife scores. *This tree is provided in `scjackknife/booster/outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre`*.


