---
title: "single cell jackknife analysis"
---

This notebook provides the workflow to calculate the single cell jackknife (scjackknife) scores for Figure 4. The scjackknife procedure randomly draws replicate cells from a larger matrix to create the 54 cell matrix. We repeat this procedure 500 times and infer a high likelihood `scjackknife tree` from each. These trees are used to calculate TBE (transfer bootstrap expectation) scores, which served as the `scjackknife scores` in Figure 4.

#Preliminaries
```{r}
#set the working directory to the analysis directory
setwd("~/repo/cellphylo/analysis")

#create directory

if(!dir.exists("scjackknife")){
  dir.create("scjackknife")
}


```

# scJackknife analysis
This workflow creates scjackknife matrices, infers a scjackknife tree from each, removes errored runs and calculates TBE scores from the scjackknife trees. 

#Create the matrices we will sample from

## Create the 919 cell 20 PC matrix (Fig. S1, step C2.1)

```{r run-pca-20pc}
pca_matrix <- run_pca(matrix_path="matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/", n_PCs=20, print=TRUE)
```


The PCA is reproducible, such that all.equal=TRUE, but it is possible for identical=FALSE when comparing PCAs performed on different machines due to very small fluctuation in values (eg due to rounding error). To facilitate reproducibility we have provided the above matrix in *`scjackknife/contml_919cell_subset_20PC_mtx`.*


## Create a 540 cell matrix to draw cell labels from 

For Figure 4, we identified and removed unstable or singleton cell type groups (Methods). Doing so results in 54 cell type groups. For the scjackknife analysis, we need to draw values from the 919 cell 20 PC matrix, but restrict the cells we sample to these 54 cell type groups. So below, we createa a 540 cell matrix (54 cell type groups, each represented by 10 replicate cells) to draw cell labels from. We do not use the values in this matrix, we just use the cell labels.
```{r create-540cell-mat}

mat <- Read10X("matrix/cross-species_integration/aqhumor_cross_species_integrated_mtx/")

black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal", "Endo", "Endo_Schlemms", "Epi-CiliaryNonPigment", "Epi-CiliaryPigment","Epi-Pigment", "MastCell", "Myoepithelium", "Neuron", "SchwalbeLine", "Uveal","Vascular","Beam-A", "Beam-X", "Epi-Corneal", "Fibroblast", "BCell")


cell_ids <- colnames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

filter_index <- which(!(cell_type_ids %in% black_list))

mat_filt <- mat[,filter_index]

#write matrix
write10xCounts("matrix_540cell_cross_species_integrated_mtx", mat_filt, version="3")
file.rename("matrix_540cell_cross_species_integrated_mtx", "scjackknife/matrix_540cell_cross_species_integrated_mtx")
```

For convenience, we provide this matrix in *`scjackknife/matrix_540cell_cross_species_integrated_mtx`*.

# Make scjack samples
To create each jackknifed matrix, we will repeatably sample 1 cell per cell type group per species from the 919 cell 20 PC matrix (Fig. S1, step C2.2), for the 54 cell type groups we used to create the phylogeny in Figure 4. We will use the above 540 cell matrix to draw cell labels from. Because PCA must be performed on the full 919 cell matrix, we will create a 919 cell 20 PC matrix (Fig. S1, step C1) and pull out the values corresponding to the cell labels we obtained from subsampling the 450 cell matrix. 

This in effect is the same as subsetting the 919 cell 20 PC matrix to 450 cell and 20 PCs (Fig. S1, step C4), but due to a quirk in the workflow the former was performed.

We do this 500 times to create 500 resampled 54 cell matrices. Each matrix is transformed into a `contml infile` and contml is used to infer  tree from it using the same settings as that used for the jumble trees.

This function will print out the 54 cell subsampled matrices as 10x feature-barcode matrices, infiles, and the list of cell ids present in each matrix.

Since the scjackknife matrices are random, we have included the input we used as *`scjackknife/input_scjack_540_TBE.tar.gz`* for reproducibility.

```{r create-scjack-sample}

# The matrix to subsample cells from. This is the 540 cell matrix (Fig. S1, step C4)
subsample_matrix_path = "scjackknife/matrix_540cell_cross_species_integrated_mtx/"
# The matrix to draw values from. This should be the 919 cell 20 PC matrix (Fig. S1, step C1)
full_matrix_path = "scjackknife/contml_919cell_subset_20PC_mtx/"
#number of scjack samples to make
n_samples = c(1:500)

#function calls: subset_combined_matrix(), run_pca(), and make_infile()
scjack_sample <- make_scjack_sample(n_samples = n_samples, subsample_matrix_path = subsample_matrix_path, full_matrix_path = full_matrix_path, file_name="540_TBE")

```
We produce 500 scjackknifed averaged matrices (`scjack_matrices/`), the `contml` infiles (`scjack_infiles/`) representing these matrices, and a list of the cell ids that were sampled for each matrix (`scjack_selected_cells/`). 

Since sampling is random, for reproducibility, we provide these input files in `scjackknife/input/input_scjack_540_TBE.tar.gz`.

It takes ~20 minutes on a 2 core 16 GiB laptop to produce 500 scjackknife samples.

# Create scripts to infer scjackknife trees with contml
Use the infiles created above as input for the `contml` runs.

`contml` was run using the same options as for the jumble trees:
  C: continuous data
  J: jumble, 100x with random seed
  G: global search

We ran this on a hpc cluster using 64 vCPUs and running 50 `contml` commands in parallel at a time. This took 4-5 hours.

Each scjack `contml` run requires its own bash script. We created and executed 500 bash scripts using the following workflow:

1. We created a template for the script that is executed for each contml run. This has been provided as `in_script_scjack_parallel_540_TBE.sh`

2. We made 500 copies of the template script. The script for this is provided as `make_reps_scjack_540_TBE.sh`.

3. Execute each scjackknife script in parallel. The script for this is `run_scjack_parallel_540_TBE.sh`.

We have included the three scripts described above in *`scripts/scjackknife'*. The scjackknife run scripts produced by running `make_reps_scjack_540_TBE.sh` has been provided in *`scripts/scjackknife/scjack_scripts.tar.gz`* for convenience.

Use the scjack infiles and the above scripts to run `contml`. When running `contml`, the scjack infiles should be in a folder called `scjack_infiles`. The scjack scripts and `run_scjack_parallel_540_TBE.sh` should be in the present working directory (ie. not in a folder).

Doing the above produced 500 high likelihood scjackknife trees. 


# Process output files
Some `contml` runs may have errored out, producing empty `outtree` files. Remove these files

Delete any empty tree files
```
find ./ -size 0 -print -delete
```

There are 439 scjackknife trees remaining.

Moving forward, use this `outtree` directory, which does not have empty tree files. This directory has been provided in *`scjackknife/output/output_processed/outtrees_scjack_540_TBE_processed`*.

Removing files causes the index between the scjack matrices and `outtree` and matrix files to be mismatched, causing downstream analysis issues. Remove those scjack matrices that produced errored runs.

```{r id-errored-runs}

#this is the raw outtree folder of scjackknife trees that has not been filtered for empty files. I do not include it but it is available upon request.
outtrees_dir = "outtrees_scjack_540_TBE/"
matrix_name_pattern = "scjackknife_540_TBE"

id_errored_runs(outtrees_dir = outtrees_dir, matrix_name_pattern = matrix_name_pattern)

```

The above function provides a list of matrices whose `contml` runs errored. This list is provided in *`scjackknife/output/output_processed/errored_runs.txt`.*

Use this list to remove these matrices using bash:
```
cd analysis/scjackknife/input_scjack_540_TBE/scjack_matrices
rm -r $(<errored_runs.txt)

```

Moving forward, use the directory of matrices with error-causing matrices removed in downstream analysis. This directory has been provided as *`scjackknife/output/output_scjack_540_TBE_processed/scjack_matrices_processed`*

# Prepare input files for Booster.

## Add common labels to scjackknife trees

scjackknife trees must share common labels (ie labels that correspond across trees) to calculate the TBE. Label the scjackknife trees with their full cell ids (for future reference) and an abbreviated cell id (common label) that can be matched across scjackknife trees.

```{r prep-scjack-boost}

n_trees =439
matrix_dir_path = "scjackknife/output/output_processed/scjack_matrices_processed/"
tree_dir_path = "scjackknife/output/output_processed/outtrees_scjack_540_TBE_processed/"
file_name_prefix = "outtrees_scjack_540_TBE"

labelled_trees <- label_scjack_trees(n_trees = n_trees, matrix_dir_path = matrix_dir_path, tree_dir_path = tree_dir_path, print=TRUE, file_name_prefix=file_name_prefix)

```

It takes ~15-16 minutes to run the above chunk on a 2 core 16 GiB laptop.

The trees with abbreviated, common labels have been provided in 
*`scjackknife/output/output_processed/outtrees_scjack_540_TBE_common_label`*, for convenience. We will use these trees to run Booster and calculate scjackknife TBE scores.


## Add common labels to focal tree

The focal tree (the tree we will plot scjackknife scores onto) must also share common labels with the scjackknife trees.
```{r prep-focal-boost}

#below is the best scoring jumble tree that used for Fig. 4
tree <- read.tree("jumble/output/outtrees_54_jumble_search_processed/outtree_54_jumble_search_1249177875.tre")


labelled_tree <- label_tree(matrix_path = "tree_inference/54_cell_tree/contml_54cell_subset_pca_var_norm_20PC_mtx/", tree_path = "jumble/output/outtrees_54_jumble_search_processed/outtree_54_jumble_search_1249177875.tre", file_name = "outtree_54_jumble_search_1249177875_label.tre", print=FALSE)

#make new labels - code stolen from scjack label
species <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[3])})
new_label <- paste0(species, "_", cluster_id)
common_labels_tree <- labelled_tree
common_labels_tree$tip.label <- new_label
#print out common labels tree
write.tree(common_labels_tree, file="outtree_54_jumble_search_1249177875_common_label.tre")



```

For convenience, this tree has been provided as *`scjackknife/booster/outtree_54_jumble_search_1249177875_common_label.tre`*

#Combine common label scjackknife trees in a single multiphylo file

Booster requires the scjackknife trees in a multi tree file.
```{r boost-files}

#combine and print out as a single tree file
multi_trees <- combine_multi_trees("scjackknife/output/output_processed/outtrees_scjack_540_TBE_common_label/")

write.tree(multi_trees, file="multiphylo_scjack_540_TBE_common_label.tre")

    
```

For convenience, this multi tree file has been provided in *`scjackknife/booster/multiphylo_scjack_540_TBE_common_label.tre`*

## Run booster to calculate TBE scores

Booster may be downloaded from `https://github.com/evolbioinfo/booster`.

```
./booster_macos64 -i outtree_54_jumble_search_1249177875_common_label.tre -b multiphylo_scjack_540_TBE_common_label.tre  > outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre
```

This produces a tree, `outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre`, with TBE scores plotted at nodes. These TBE scores were used in Fig. 4 as the scjackknife scores. This tree is provided in *`scjackknife/booster/outtree_54_jumble_search_1249177875_scored_with_scjack540TBE.tre`*.


