---
title: "average"
editor_options: 
  chunk_output_type: console
---

This notebook walks you through the steps to create Figure 5. This tree is inferred from averaged values - tips represent a set of averaged cells, rather than a single cell representative, for the cell type group.

# Preliminaries
```{r}
setwd("~/repo/cellphylo/analysis")

library(Seurat)
library(tidyverse)
library(DropletUtils)
library(phangorn)
```

### Create a list of cells to draw from

Create a list of cells to sample cell ids from. The 919 cell matrix has all 92 cell type groups. As in Figure 4, we remove unstable or singleton cell types, leaving 530 cells representing 53 cell type groups. In this case, we restrict sampled cell ids to the 540 cells from 54 cell type groups (the same cell type groups used in Figure 4). 

```{r permissible_cells}
library(Seurat)

#identical=TRUE to original average 919 cell 20 PC matrix
mat <- Read10X("scjackknife/contml_919cell_subset_20PC_mtx/")

#54 cell type groups
black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal" , "Endo", "Epi-CiliaryNonPigment","Epi-CiliaryPigment" , "Epi-Pigment" , "MastCell" ,  "Myoepithelium" , "Neuron" ,  "SchwalbeLine", "Uveal", "Vascular", "Fibroblast", "Epi-Corneal", "Beam-X", "Beam-A", "BCell")

cell_ids <- rownames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

#filter matrix
filter_index <- which(!(cell_type_ids %in% black_list))
mat_filt <- mat[filter_index,]
permissible_cells <- rownames(mat_filt)

#print out the file for later use
write.table(permissible_cells, file="540_cell_subset_cell_ids.txt", sep="\t", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

# Infer averaged tree and calculate Jumble scores
The 919 cell 20 PC matrix (Fig. S1, step C2.1) has ~10 replicate cells per cell type group, with 92 cell type groups represented, and 20 PCs. We average these 10 cells for each cell type group, removing unstable cell type groups or singletons, to produce a 53 averaged-cell 20 PC matrix (Fig. S1, Step D1.2). This matrix represents 53 cell type groups, each represented by the averaged value of its 10 replicate cells (Methods).

## Create the 54 averaged-cell 20 PC matrix (Fig. S1, step D1.2)

We first take the 919 cell 20 PC matrix (Fig. S1, step C2.1) and average each of its ten replicates to produce a 92 averaged-cell 20 PC matrix (Fig. S1, step D1.1).

```{r make_pca_mean_mat}
library(Seurat)
library(tidyverse)
library(DropletUtils)

#Input matrix
#919 cells - 10 cells per cell type cluster, 92 cell type clusters
mat <- Read10X("scjackknife/contml_919cell_subset_20PC_mtx")
mat <- t(as.matrix(mat))

permissible_cells <- read.table("average/540_cell_subset_cell_ids.txt") %>% unlist()
mat.filtered <- mat[,which(colnames(mat) %in% permissible_cells)]

#parse out annotations
orig_ids <- colnames(mat.filtered)
species <- sapply(strsplit(orig_ids, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(orig_ids, "_"), function(v){return(v[3])})
cell_type_id <- sapply(strsplit(orig_ids, "_"), function(v){return(v[4])})
species_cluster <- paste0(species, "_", cluster_id)
#54 species cluster ids
unique.species_cluster <- unique(species_cluster)


means <- sapply(1:length(unique.species_cluster), function(i){
  

  mat.subset <- mat.filtered[,which(species_cluster == unique.species_cluster[i])]
  means.df <- rowMeans(mat.subset) %>% as.data.frame()
  colnames(means.df) <- unique.species_cluster[i]
  
  return(means.df)
  
}) #close apply

means_mat <- as.data.frame(means)
#dashes had been replaced with ".". Not sure why, make dashes again

fixed_names  <- gsub("\\.", "-", colnames(means_mat))
species <- sapply(strsplit(fixed_names, "_"), function(v){return(v[1])})
cluster <-sapply(strsplit(fixed_names, "_"), function(v){return(v[2])})
cell_type <- gsub('[0-9+]', "", cluster)
new_names <- paste0(species, "_sample_", cluster, "_", cell_type, "_barcode")
colnames(means_mat) <- new_names
rownames(means_mat) <- rownames(mat)

#return back to post PCA orientation
means_mat <- t(means_mat)

means_mat.sparse <- as.sparse(means_mat)
write10xCounts("contml_54means_20PC_afterPCA_mtx", means_mat.sparse, version="3")

```

## Create infile for contml
```{r}
make_infile(matrix_path = "contml_54means_20PC_afterPCA_mtx", print=TRUE)
```

## Infer Jumble trees using contml
Infer the jumble trees from the infile for the 53 averaged-cell 20 PC matrix (Fig. S1, step D1.2) (ie from `contml_53mean_pca_var_norm_20PC_afterPCA_mtx`).

The bash script `script_jumble_parallel_54mean_20PC_afterPCA.sh` was used to run contml. The script `run_jumble_parallel_54mean_20PC_afterPCA.sh` was used to run that script in parallel on a hpc cluster. *These scripts are provided in `analysis/scripts/average/jumble/`.*

The `contml` settings:
  C: continuous characters
  J:   Randomize input order of species (jumble)
    Random number to seed (see bash script)
    Number of times to jumble: 100

## Remove missing/errored files
Some Jumble runs may error and produce no tree. We must remove those files.

Delete any empty tree files using the bash command.
```
find ./ -size 0 -print -delete
```
This leaves 209 trees.

The processed outtree files have been provided in *`average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed.tar.gz`*.

## Identify best jumble tree
```{r}
library(phangorn)

tree_dir_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed"

#find jumble tree with highest sum of scores
best_jumble_tree(tree_dir_path = tree_dir_path)
list.files(tree_dir_path)

```

5 trees are tied for highest sum of scores.
outtree_54mean_20PC_afterPCA_jumble_search_1687210069.tre
outtree_54mean_20PC_afterPCA_jumble_search_1910156897.tre
outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre
outtree_54mean_20PC_afterPCA_jumble_search_2741266717.tre
outtree_54mean_20PC_afterPCA_jumble_search_3594350545.tre

Choose the "best" best jumble tree that is most repesentative topologically of the 5 "best" jumble trees. We calculated Robinson-Foulds, path distance, and Kuhner-Felsenstein distance.


```{r}
library(phangorn)

#create a multiphylo of best trees
tree_dir_path <- "average/jumble/output/best_trees/"
multiphylo <- combine_multi_trees(tree_dir_path)

#make a distance matrix. See phangorn::treedist for description of distance metrics.
#using Robinson-Foulds distance
RF_dist_mtx <- RF.dist(multiphylo) 
RF_dist_mat <- as.matrix(RF_dist_mtx)
colsums <- colSums(RF_dist_mat)
min <- min(colsums)
RF_dist_mat[,which(colSums(RF_dist_mat) == min)] #all 0

#using path distance
path_dist_mtx <- path.dist(multiphylo) 
path_dist_mat <- as.matrix(path_dist_mtx)
colsums <- colSums(path_dist_mat)
min <- min(colsums)
path_dist_mat[,which(colSums(path_dist_mat) == min)] # all 0

#using Kuhner & Felsenstein (1994) distance.
KF_dist_mtx <- KF.dist(multiphylo) 
KF_dist_mat <- as.matrix(KF_dist_mtx)
colsums <- colSums(KF_dist_mat)
min <- min(colsums)
KF_dist_mat[,which(colSums(KF_dist_mat) == min)] # tree 3 has shortest distance


list.files(tree_dir_path)

```

The most representative, highest-scoring jumble tree is: `outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre`. It was selected as the focal phylogeny for Figure 5.


Score and label best jumble tree:
```{r}
#make multiphylo object of jumble trees
tree_dir_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/"
multiphylo <- combine_multi_trees(tree_dir_path)

#focal tree (tree to plot jumble scores)
focal_tree_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre"

tree <- score_jumble(multiphylo = multiphylo, focal_tree_path = focal_tree_path, print=TRUE)

#label tree
mat_path <- "contml_54means_20PC_afterPCA_mtx/"
tree_path <- "scored_tree.tre"
file_name <- "54mean_20PC_afterPCA_jumble_search_2525674179_scored_label.tre"
label_tree(mat_path, tree_path, file_name, print=TRUE)

```

*The file `54mean_20PC_afterPCA_jumble_search_2525674179_scored_label.tre` was used as the focal tree in Figure 5. It is provided in `average/jumble/output/`.*


# Calculate scJackknife scores
To perform the scJackknife analysis, we take the 919 cell 20 PC matrix (Fig. S1, step C2.1), which has ~10 cells per cell type group (92 cell type groups) and randomly sample and average 5 of those 10 cells for each cell type group.

## Make scJackknife input files

### Make scjack samples
As in the scJackknife tutorial (`5_scJackknife_analysis.Rmd`), we create scjack samples by randomly sampling 5 cells per cell type group from the 919 cell 20 PC matrix (Fig. S1, step C1.1), limiting our cell type groups to the 54 described above. 

The below code creates  `contml` infiles, matrices, and a list of cells that were sampled to make the averaged matrices.

```{r}
setwd("~/repo/cellphylo/analysis")

#parameters
n_samples = 500
matrix_path = "scjackknife/contml_919cell_subset_20PC_mtx"
sample_size=5
print=TRUE
file_name = "540_cells_20PC_after_PCA"
# limit cells to the 540 cells that represent the filtered 54 cell type groups
limit_cells <- read.table("average/540_cell_subset_cell_ids.txt")


#run function
make_scjack_sample_mean(n_samples=n_samples, mat=matrix_path, sample_size=sample_size, limit_cells=limit_cells, file_name=file_name, print=print)

```


We produce 500 scjackknifed averaged matrices, the `contml` infiles representing these matrices, and a list of the 5 cells sampled for each cell type group for each matrix.

We provide these input files in *`average/scjackknife/input/input_540_cells_20PC_after_PCA.tar.gz`.*

### Create scripts to infer scjackknife trees with contml
Use the infiles created above to run `contml`.

`contml` was run using the same options as for the jumble trees:
  C: continuous data
  J: jumble, 100x with random seed

1. We created a template for the script that is executed for each contml run. This has been provided as `in_script_scjack_parallel_means_5reps_540.sh`

2. We made 500 copies of the template script. The script for this is provided as `make_reps_scjack_means_5reps_540.sh`.

3. Execute each scjackknife script in parallel. The script for this is `run_scjack_parallel_means_5reps_540.sh`.
infiles should be in a directory called `scjack_infiles`
scripts should be present in the working directory (ie not in a directory)

We have included the three scripts described above in *`scripts/average/scjack'* and the scjackknife run scripts (the output of running the three above scripts) in *`scripts/scjackknife/scjack_scripts_mean.tar.gz`*

Using these scripts to run `contml` produced 500 outtrees representing a scjackknife tree inferred from each 54 averaged cell matrix.


## Process output files
Some `contml` runs may have errored out, producing empty `outtree` files. Remove these files

Delete any empty tree files
```
find ./ -size 0 -print -delete
```
393/500 trees survived.

Moving forward, use the directories of `outtrees` that have had errored runs removed for downstream analyses. *This directory has been provided as `average/scjackknife/output/output_processed/outtrees_scjack_means_5reps_540_processed.tar.gz`.*

Removing files causes the index between the scjack matrices and `outtree` and matrix files to be mismatched, causing downstream analysis issues. Remove those scjack matrices that produced errored runs.

Identify runs that errored:
```{r}

outtrees_dir = "~/repo/aqhumor/AqHum_5Species/aqhumor_final/average/trees_ave/scjack/after_PCA/540_cells/output_540_cells_20PC_after_PCA/outtrees_scjack_means_5reps_540/"
matrix_name_pattern = "means_5_reps_540_cells_20PC_after_PCA"

id_errored_runs_mean(outtrees_dir = outtrees_dir, matrix_name_pattern = matrix_name_pattern)

```
The above function provides a list of matrices whose `contml` runs errored. This list is provided in *`average/scjackknife/output/output_processed/errored_runs_mean.txt`.*

Remove corresponding matrices
```
cd ~/repo/aqhumor/AqHum_5Species/aqhumor_final/average/trees_ave/scjack/after_PCA/540_cells/input_540_cells_20PC_after_PCA/scjack_matrices
rm -r $(<errored_runs_mean.txt)

```

Moving forward, use the directory of matrices with error-causing matrices removed in downstream analysis. *This has been provided as `average/scjackknife/output/output_processed/scjack_matrices_processed.tar.gz`*


## Prepare input files for Booster.

### Add common labels to scjackknife trees
scjackknife trees must share common labels in order to calculate the TBE. Label the scjackknife trees with their full cell ids and an abbreviated cell id that can be matched across scjackknife trees.

```{r prep-scjack-boost}

n_trees = 393
matrix_dir_path = "average/scjackknife/output/output_processed/scjack_matrices_processed/"
tree_dir_path = "average/scjackknife/output/output_processed/outtrees_scjack_means_5reps_540_processed/"
file_name_prefix = "outtree_scjack_means_5reps_540"

labelled_trees <- label_scjack_trees(n_trees = n_trees, matrix_dir_path = matrix_dir_path, tree_dir_path = tree_dir_path, print=TRUE, file_name_prefix=file_name_prefix)

```
The trees that have been labelled with common labels have been provided in 
*`average/scjackknife/output/output_processed/outtree_scjack_means_5reps_540_common_label.tar.gz`*

### Add common labels to best Jumble tree

```{r}

tree <- read.tree("average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre")

labelled_tree <- label_tree(matrix_path = "../../aqhumor/AqHum_5Species/aqhumor_final/average/trees_ave/jumble/54_mean/54mean_20PC_afterPCA/contml_54means_20PC_afterPCA_mtx/", tree_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre", file_name = "outtree_54mean_20PC_afterPCA_jumble_search_2525674179_label.tre", print=FALSE)


#make new labels - code stolen from scjack label
species <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[3])})
new_label <- paste0(species, "_", cluster_id)
common_labels_tree <- labelled_tree
common_labels_tree$tip.label <- new_label
#print out common labels tree
write.tree(common_labels_tree, file="outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre")

```

To facilitate reproduction, this tree has been provided in *`average/booster/outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre`*


### Combine scJackknife trees into a multiphylo .tre file for Booster

```{r boost-files}

#combine and print out as a single tree file
multi_trees <- combine_multi_trees("average/scjackknife/output/output_processed/outtree_scjack_means_5reps_540_common_label")

write.tree(multi_trees, file="multiphylo_scjack_means_5reps_540_common_label.tre")

```
To facilitate reproduction, this multiphylo tree file has been provided in *`average/booster/multiphylo_scjack_means_5reps_540_common_label.tre`*

## Run booster

```
./booster_macos64  -i outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre -b multiphylo_scjack_means_5reps_540_common_label.tre > outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label_best_tree_TBE.tre
```

This produces a tree, `outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label_best_tree_TBE.tre`, with TBE scores plotted at nodes. These TBE scores were used in Figure 5 as the scjackknife scores. *This tree is provided in 'average/booster`*.

