---
title: "average"
editor_options: 
  chunk_output_type: console
---

This notebook walks you through the steps to create Figure 5. This tree is inferred from averaged values - tips represent a set of averaged cells, rather than a single cell representative, for the cell type group.

# Preliminaries
```{r}
#set the working directory to the analysis directory
setwd("~/repo/cellphylo/analysis")

```

# Create a list of cells to draw from
The following workflow draws values from the 919 cell 20 PC matrix (Fig. S1, step C2.1) but limits the cell type groups to the 54 cell type groups featured in Figure 4. Here, we make a list of cells from the 919 cell matrix that only includes cell ids from the 540 cell type groups. This list has 540 cells, with each of the 54 cell type group represented by 10 cells. We will use this list to restrict downstream analyses.

```{r permissible_cells}

#identical=TRUE to original average 919 cell 20 PC matrix
mat <- Read10X("scjackknife/contml_919cell_subset_20PC_mtx/")

#54 cell type groups
black_list <- c("Beam-B", "Beam-Y", "CollectorChnlAqVein", "Corneal" , "Endo", "Epi-CiliaryNonPigment","Epi-CiliaryPigment" , "Epi-Pigment" , "MastCell" ,  "Myoepithelium" , "Neuron" ,  "SchwalbeLine", "Uveal", "Vascular", "Fibroblast", "Epi-Corneal", "Beam-X", "Beam-A", "BCell")

cell_ids <- rownames(mat)
cell_type_ids <- sapply(strsplit(cell_ids, "_"), function(v){return(v[4])})

#filter matrix
filter_index <- which(!(cell_type_ids %in% black_list))
mat_filt <- mat[filter_index,]
permissible_cells <- rownames(mat_filt)

#print out the file for later use
write.table(permissible_cells, file="540_cell_subset_cell_ids.txt", sep="\t", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

We will restrict our following analyses to this 540 cel subset of the 919 cell 20 PC matrix.


# Infer Figure 5 focal tree and calculate Jumble scores

The 919 cell 20 PC matrix (Fig. S1, step C2.1) has ~10 replicate cells per cell type group, with 92 cell type groups represented, and 20 PCs. We average these 10 cells for each cell type group, removing unstable cell type groups or singletons, to produce a 54 averaged-cell 20 PC matrix (Fig. S1, Step D1). This matrix represents 54 cell type groups, each represented by the averaged value of its 10 replicate cells (Methods).

## Create the 54 averaged-cell 20 PC matrix (Fig. S1, step D1)

We first take the 919 cell 20 PC matrix (Fig. S1, step C2.1) and average each of its ten replicates, selecting the 54 cell type groups in our permissible cells list. This creates a 54 averaged-cell 20 PC matrix (Fig. S1, step D1)

```{r make_pca_mean_mat}

#Input matrix
#919 cells - 10 cells per cell type cluster, 92 cell type clusters
mat <- Read10X("scjackknife/contml_919cell_subset_20PC_mtx")
mat <- t(as.matrix(mat))

#limit the cells to the 54 cell type groups
permissible_cells <- read.table("average/540_cell_subset_cell_ids.txt") %>% unlist()
mat.filtered <- mat[,which(colnames(mat) %in% permissible_cells)]

#parse out annotations
orig_ids <- colnames(mat.filtered)
species <- sapply(strsplit(orig_ids, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(orig_ids, "_"), function(v){return(v[3])})
cell_type_id <- sapply(strsplit(orig_ids, "_"), function(v){return(v[4])})
species_cluster <- paste0(species, "_", cluster_id)
#54 species cluster ids
unique.species_cluster <- unique(species_cluster)


means <- sapply(1:length(unique.species_cluster), function(i){
  

  mat.subset <- mat.filtered[,which(species_cluster == unique.species_cluster[i])]
  means.df <- rowMeans(mat.subset) %>% as.data.frame()
  colnames(means.df) <- unique.species_cluster[i]
  
  return(means.df)
  
}) #close apply

means_mat <- as.data.frame(means)
#dashes had been replaced with ".". Not sure why, make dashes again

fixed_names  <- gsub("\\.", "-", colnames(means_mat))
species <- sapply(strsplit(fixed_names, "_"), function(v){return(v[1])})
cluster <-sapply(strsplit(fixed_names, "_"), function(v){return(v[2])})
cell_type <- gsub('[0-9+]', "", cluster)
new_names <- paste0(species, "_sample_", cluster, "_", cell_type, "_barcode")
colnames(means_mat) <- new_names
rownames(means_mat) <- rownames(mat)

#return back to post PCA orientation
means_mat <- t(means_mat)

means_mat.sparse <- as.sparse(means_mat)
write10xCounts("contml_54means_20PC_afterPCA_mtx", means_mat.sparse, version="3")

```


## Create infile for contml

Create the infile to run `contml` on from this 54 averaged-cell 20 PC matrix.
```{r}
make_infile(matrix_path = "contml_54means_20PC_afterPCA_mtx", print=TRUE)
```

This infile has been provided in *`average/jumble/input/infile_54means_20PC_afterPCA`.* `contml ` expects the input file to be named "infile", so rename before use.

## Infer Jumble trees using contml

Infer the jumble trees from the infile for the 54 averaged-cell 20 PC matrix (Fig. S1, step D1).

The bash script `script_jumble_parallel_54mean_20PC_afterPCA.sh` was used to run contml. The script `run_jumble_parallel_54mean_20PC_afterPCA.sh` was used to run that script in parallel on a hpc cluster. *These scripts are provided in `analysis/scripts/average/jumble/`.*

The `contml` settings:
  C: continuous characters
  J: Randomize input order of species (jumble)
      Random number to seed (see bash script)
      Number of times to jumble: 100

## Remove missing/errored files
Some Jumble runs may error and produce no tree. We must remove those files.

Delete any empty tree files using the bash command.
```
find ./ -size 0 -print -delete
```
This leaves 209 trees.

The processed outtree files have been provided in *`average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed.tar.gz`*.

## Identify best jumble tree
```{r}

tree_dir_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed"

#find jumble tree with highest sum of scores
best_jumble_tree(tree_dir_path = tree_dir_path)
list.files(tree_dir_path)

```

5 trees are tied for highest sum of scores.
outtree_54mean_20PC_afterPCA_jumble_search_1687210069.tre
outtree_54mean_20PC_afterPCA_jumble_search_1910156897.tre
outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre
outtree_54mean_20PC_afterPCA_jumble_search_2741266717.tre
outtree_54mean_20PC_afterPCA_jumble_search_3594350545.tre

Deposit these trees in a directory called `best_trees` in `average/jumble/output/`. This directory is not provided but these trees are available in `average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed`.

Choose the "best" best jumble tree that is most repesentative topologically of the 5 "best" jumble trees. We calculated Robinson-Foulds, path distance, and Kuhner-Felsenstein distance.


```{r}

#create a multiphylo of best trees
tree_dir_path <- "average/jumble/output/best_trees/"
multiphylo <- combine_multi_trees(tree_dir_path)

#make a distance matrix. See phangorn::treedist for description of distance metrics.
#using Robinson-Foulds distance
RF_dist_mtx <- RF.dist(multiphylo) 
RF_dist_mat <- as.matrix(RF_dist_mtx)
colsums <- colSums(RF_dist_mat)
min <- min(colsums)
RF_dist_mat[,which(colSums(RF_dist_mat) == min)] #all 0

#using path distance
path_dist_mtx <- path.dist(multiphylo) 
path_dist_mat <- as.matrix(path_dist_mtx)
colsums <- colSums(path_dist_mat)
min <- min(colsums)
path_dist_mat[,which(colSums(path_dist_mat) == min)] # all 0

#using Kuhner & Felsenstein (1994) distance.
KF_dist_mtx <- KF.dist(multiphylo) 
KF_dist_mat <- as.matrix(KF_dist_mtx)
colsums <- colSums(KF_dist_mat)
min <- min(colsums)
KF_dist_mat[,which(colSums(KF_dist_mat) == min)] # tree 3 has shortest distance


list.files(tree_dir_path)

```

The most representative, highest-scoring jumble tree is: `outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre`. It was selected as the focal phylogeny for Figure 5.


Score and label best jumble tree:
```{r}

#make multiphylo object of jumble trees
tree_dir_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/"
multiphylo <- combine_multi_trees(tree_dir_path)

#focal tree (tree to plot jumble scores)
focal_tree_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre"

tree <- score_jumble(multiphylo = multiphylo, focal_tree_path = focal_tree_path, print=TRUE)

#label tree
#this matrix was produced is Fig. S1 step D1. It is not provided as you may produce it above.
mat_path <- "contml_54means_20PC_afterPCA_mtx/"
tree_path <- "scored_tree.tre"
file_name <- "54mean_20PC_afterPCA_jumble_search_2525674179_scored_label2.tre"
label_tree(mat_path, tree_path, file_name, print=TRUE)

```

*The file `54mean_20PC_afterPCA_jumble_search_2525674179_scored_label.tre` was used as the focal tree in Figure 5. It is provided in `average/jumble/output/`.*


# Calculate scJackknife scores
To perform the scJackknife analysis, we take the 919 cell 20 PC matrix (Fig. S1, step C2.1), which has ~10 cells per cell type group (92 cell type groups) and randomly sample and average 5 of those 10 cells to produce a jackknifed matrix of 54 averaged-cells and 20 PCs (Fig. S1, step D2).


## Make scJackknife input files

### Make scjack matrices (Fig. S1, step D2)
We create scjackknife samples by randomly sampling 5 cells per cell type group from the 919 cell 20 PC matrix (Fig. S1, step C2.1), limiting our cell type groups to the 54 described above. Each jackknifed matrix has 54 values, each representing the average of five randomly chosen replicates, and 20 PCs (Fig. S1, step D2). We do this 500 times to create 500 scjackknife matrices.

The below code creates  `contml` infiles, matrices, and a list of cells that were sampled to make the averaged matrices.

```{r}

#parameters
n_samples = 500
matrix_path = "scjackknife/contml_919cell_subset_20PC_mtx"
sample_size=5
print=TRUE
file_name = "540_cells_20PC_after_PCA2"
# limit cells to the 540 cells that represent the filtered 54 cell type groups
limit_cells <- read.table("average/540_cell_subset_cell_ids.txt")


#run function
make_scjack_sample_mean(n_samples=n_samples, mat=matrix_path, sample_size=sample_size, limit_cells=limit_cells, file_name=file_name, print=print)

```


We produce 500 scjackknifed averaged matrices (`scjack_matrices/`), the `contml` infiles (`scjack_infiles/`) representing these matrices, a list of cell ids in the order they are found in the matrix (`scjack_matrix_cell_ids/`), and a list of the 5 cells sampled for each cell type group for each matrix (`scjack_selected_cells/`). 

This takes ~8-10 min on a laptop with 2 cores and 16 GiB memory.

For reproducibility, we provide these input files in *`average/scjackknife/input/input_540_cells_20PC_after_PCA.tar.gz`.*

### Create scripts to infer scjackknife trees with contml
Use the infiles created above to run `contml`.

`contml` was run using the same options as for the jumble trees:
  C: continuous characters
  J: Randomize input order of species (jumble)
      Random number to seed (see bash script)
      Number of times to jumble: 100

1. We created a template for the script that is executed for each contml run. This has been provided as `in_script_scjack_parallel_means_5reps_540.sh`

2. We made 500 copies of the template script. The script for this is provided as `make_reps_scjack_means_5reps_540.sh`.

3. Execute each scjackknife script in parallel. The script for this is `run_scjack_parallel_means_5reps_540.sh`.
infiles should be in a directory called `scjack_infiles`
scripts should be present in the working directory (ie not in a directory)

We have included the three scripts described above in *`scripts/average/scjack'* and the scjackknife run scripts (the output of `make_reps_scjack_means_5reps_540.sh`) in *`scripts/scjackknife/scjack_scripts_mean.tar.gz`*

Using these scripts to run `contml` produced 500 scjackknife trees.


## Process output files
Some `contml` runs may have errored out, producing empty `outtree` files. Remove these files

Delete any empty tree files
```
find ./ -size 0 -print -delete
```
393/500 trees survived.

Moving forward, use the directories of `outtrees` that have had errored runs removed for downstream analyses. This directory has been provided as *`average/scjackknife/output/output_processed/outtrees_scjack_means_5reps_540_processed.tar.gz`.*

Removing files causes the index between the scjack matrices and `outtree` files to be mismatched, causing downstream analysis issues. Remove those scjack matrices that produced errored runs.

Identify runs that errored:
```{r}

#this is the outtree folder of scjackknife trees that has not been filtered for empty files. I do not include it but it is available upon request. 
outtrees_dir = "outtrees_scjack_means_5reps_540/"
matrix_name_pattern = "means_5_reps_540_cells_20PC_after_PCA"

id_errored_runs_mean(outtrees_dir = outtrees_dir, matrix_name_pattern = matrix_name_pattern)

```
The above function provides a list of matrices whose `contml` runs errored. This list is provided in *`average/scjackknife/output/output_processed/errored_runs_mean.txt`.*

Remove corresponding matrices
```
cd ~/repo/aqhumor/AqHum_5Species/aqhumor_final/average/trees_ave/scjack/after_PCA/540_cells/input_540_cells_20PC_after_PCA/scjack_matrices
rm -r $(<errored_runs_mean.txt)

```

Moving forward, use the directory of matrices with error-causing matrices removed in downstream analysis. This has been provided as *`average/scjackknife/output/output_processed/scjack_matrices_processed.tar.gz`*


## Prepare input files for Booster.

### Add common labels to scjackknife trees
scjackknife trees must share common labels in order to calculate the TBE. Label the scjackknife trees with their full cell ids and an abbreviated cell id that can be matched across scjackknife trees.

```{r prep-scjack-boost}

n_trees = 393
matrix_dir_path = "average/scjackknife/output/output_processed/scjack_matrices_processed/"
tree_dir_path = "average/scjackknife/output/output_processed/outtrees_scjack_means_5reps_540_processed/"
file_name_prefix = "outtree_scjack_means_5reps_540"

labelled_trees <- label_scjack_trees(n_trees = n_trees, matrix_dir_path = matrix_dir_path, tree_dir_path = tree_dir_path, print=TRUE, file_name_prefix=file_name_prefix)

```
This takes ~8-10 min on a laptop with 2 cores and 16 GiB memory.

The trees that have been labelled with common labels have been provided in 
*`average/scjackknife/output/output_processed/outtree_scjack_means_5reps_540_common_label.tar.gz`*

### Add common labels to best Jumble tree

```{r}

tree <- read.tree("average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre")

#contml_54means_20PC_afterPCA_mtx/ is not provided but you may generate it as described above.
labelled_tree <- label_tree(matrix_path = "../../aqhumor/AqHum_5Species/aqhumor_final/average/trees_ave/jumble/54_mean/54mean_20PC_afterPCA/contml_54means_20PC_afterPCA_mtx/", tree_path = "average/jumble/output/outtrees_54mean_20PC_afterPCA_jumble_search_processed/outtree_54mean_20PC_afterPCA_jumble_search_2525674179.tre", file_name = "outtree_54mean_20PC_afterPCA_jumble_search_2525674179_label.tre", print=FALSE)


#make new labels - code stolen from scjack label
species <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[1])})
cluster_id <- sapply(strsplit(labelled_tree$tip.label, "_"), function(v){return(v[3])})
new_label <- paste0(species, "_", cluster_id)
common_labels_tree <- labelled_tree
common_labels_tree$tip.label <- new_label
#print out common labels tree
write.tree(common_labels_tree, file="outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre")

```

To facilitate reproduction, this tree has been provided in *`average/booster/outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre`*


### Combine scJackknife trees into a multiphylo .tre file for Booster

```{r boost-files}

#combine and print out as a single tree file
multi_trees <- combine_multi_trees("average/scjackknife/output/output_processed/outtree_scjack_means_5reps_540_common_label")

write.tree(multi_trees, file="multiphylo_scjack_means_5reps_540_common_label.tre")

```
To facilitate reproduction, this multiphylo tree file has been provided in *`average/booster/multiphylo_scjack_means_5reps_540_common_label.tre`*

## Run booster

Booster may be downloaded from `https://github.com/evolbioinfo/booster`.

```
./booster_macos64  -i outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label.tre -b multiphylo_scjack_means_5reps_540_common_label.tre > outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label_best_tree_TBE.tre
```

This produces a tree, `outtree_54mean_20PC_afterPCA_jumble_search_2525674179_common_label_best_tree_TBE.tre`, with TBE scores plotted at nodes. These TBE scores were used in Figure 5 as the scjackknife scores. *This tree is provided in 'average/booster`*.

